Deriving Random Generators via\\ Higher-Level Representations

### 2019-01-20 paper/sections.old/hrep.tex
WIP

The key idea of this work is to represent different sources of structured value
information of a data type in a homogeneous way that we call a ``higher-level
representation'' (|HRep| from now on).%
\footnote{The notion of higher-level comes from that the generation process of
  given target data type is entirely determined by the type of the chosen
  representation, instead of by a concrete generator defined at the term level.}
%
For this purpose, we use a series of automatically derived data types, each one
representing an atomic unit of information that can be randomly generated and
then reflected back to the corresponding value of the original data type.
%
Later, the user can compose these atomic representations using the provided type
level combinators in different ways into a ``generation specification'' that
completely determines the generation process.

@@

Perhaps more interesting, we define a combinator $(\oplus)$ to compose two
|HRep|s into a single one using a sum type and representing a random choice
between them:

@@

A composite representation built using $(\oplus)$ is transformed back into the
target data type by pattern matching on the data type variant and evaluating the
inner |HRep|:
%
\begin{alignat*}{3}
  &\evalrep{L\ x &&: f \oplus g}{t} = \evalrep{x : f &&}{t} \\
  &\evalrep{R\ x &&: f \oplus g}{t} = \evalrep{y : g &&}{t}
\end{alignat*}

@@

The previous definition we reflect the type level frequencies of |f| and |g|
(|freq at_f| and |freq at_g|).
%
This frequency reflection defaults to $1$ if no frequency tag ($\otimes$) is
used.
%
Then we use these frequencies to generate each inner |HRep| in the apropriate
proportion.
%
When the generation size gets exhausted, we reflect the terminal generation
frequency of each inner |HRep| in the same way as before (|freq0 at_f| and
|freq0 at_g|).
%
This time, however, we default the frequency reflection to $0$ for any inner
that not tagged as terminal, avoiding to generate non-terminal constructions in
the last step.

With the introduced combinators, we can easily create a type synonym |HRep_Exp|
to specify a generation schema equivalent to the one seen in the concrete random
generator of type |Exp| presented in Section \ref{sec:randomtesting}:

@@

However, a value of type |HRep_Exp| still has its recursive calls abstracted
away---the type parameter |r| is implicit in the definition of |HRep_Exp|.
%
If we want to generate recursive values using this representation we need to
introduce a last type level combinator to ``tie the knot'' recursively:

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation |f|, we do not parametrize the generation of the recursive fields
of the inner represention over an external generator |gen_r|.
%
Instead, we can tie the recursive knot in our random generator by piping the
generation of the inner |HRep|, and calling itself to generate recursive values:

@@

This way we obtain a concrete recursive generator for each representation |f|
that we define.
%
With this generator, we can define a random generator for our target data type
simply by evaluating a random value of the representation:

@@

So far we have introduced the representation machinery required to represent the
random generation of a target data type considering only the structure encoded
on its definition.
%
This approach has the advantage of being extendable with different sources of
structured information.
%
We proceed to introduce two extensions that help to address the problematic
testing scenarios presented in the previous section.

@@

The first pattern of |foo| conteins two pattern variables (|x| and |y|) of type
|Exp| that are abstracted in the fields of |Mk_foo_1|.
%
Similarly, the second pattern of |foo| contains the pattern variable |x| of type
|Int| represented by the first field of |Mk_foo_2|, along with the pattern
variable |y| of type |Exp| that also is abstracted in the second field of
|Mk_foo_2|.

To evaluate these representations back to our target data type, we simply expand
them it into the concrete value represented by the original pattern:
%
\begin{align*}
  &\evalrep{|Mk_foo_1|\ x\ y}{Exp} = \\
  &\ \ |Add (Add eval_x_Exp (Val 50)) (Add (Val 25) eval_y_Exp)|\\
  &\evalrep{|Mk_foo_2|\ x\ y}{Exp} = \\
  &\ \ |Mul (Val 50) (Mul (Val x) eval_y_Exp)|
\end{align*}

And the random generation of pattern |HRep|s is defined in the same way as we
did before with data constructors |HRep|s:

@@

Moreover, the generation procedure for abstract interface representations
follows the same pattern as before:

@@

Finally, we can put all the derived machinery together into a generation
specification |Spec| for the type |Exp|, assigning (possibly) different
generation frequencies to each individual |HRep| that constitutes it:

\begin{code}
type Spec  =       HRep_Exp  otimes 4
           oplus'  HRep_foo  otimes 2
           oplus'  HRep_M
\end{code}

@@

This representation can be extended to mutually recursive and parametric types
...

@@

>>>>

The key idea of this work is to represent different sources of structured value
information of a data type in a homogeneous way that we call a ``higher-level
representation'' (|HRep| from now on).%
\footnote{The notion of higher-level comes from that the generation process of
  given target data type is entirely determined by the type of the chosen type
  level representation, instead of by a concrete generator defined at the term
  level.}
%
For this purpose, we use a series of automatically derived data types, each one
representing an atomic unit of information that can be randomly generated and
then reflected back to the corresponding value of the original data type.
%
Later, the user can compose these atomic representations using the provided type
level combinators in different ways into a ``generation specification'' that
completely determines the generation process.

@@

Perhaps more interesting, we define a combinator $(\oplus)$ to compose two
|HRep|s into a single one using a sum type to represent a random choice between
them:

@@

A composite representation built using $(\oplus)$ is transformed back into the
target data type by pattern matching on the data type variant and evaluating the
inner |HRep| accodingly:
%
\begin{alignat*}{3}
  &\evalrep{L\ x &&: f \oplus g}{t} = \evalrep{x : f &&}{t} \\
  &\evalrep{R\ x &&: f \oplus g}{t} = \evalrep{y : g &&}{t}
\end{alignat*}

@@

In the previous definition, we reflect the type level frequencies of |f| and |g|
(|freq at_f| and |freq at_g|).
%
This reflection defaults to $1$ if the frequency tag ($\otimes$) is not present.
%
Then we use these frequencies to generate each inner |HRep| in the apropriate
proportion.
%
When the generation size gets exhausted, we reflect the terminal generation
frequency of each inner |HRep| in the same way as before (|freq0 at_f| and
|freq0 at_g|).
%
This time, however, we default the frequency reflection to $0$ for any inner
that not tagged as terminal, avoiding to generate non-terminal constructions in
the last step.

Later, we can use these combinators to create a type synonym |HRep_Exp| that
specifies a generation schema equivalent to the one seen in the concrete random
generator of type |Exp| presented in Section \ref{sec:randomtesting}:

@@

However, a value of type |HRep_Exp| still has its recursive calls abstracted
away---the type parameter |r| is implicit at the definition of |HRep_Exp|.
%
We can think of it as a ``single layer'' of representation.
%
To make it able to represent recursive values we need to define a last type
level combinator to ``tie the knot'':

@@

Unlike the other combinators, to randomly generate a fixed point of a certain
representation |f|, we do not needt to parametrize the generation of the
recursive fields of the inner represention over an external generator |gen_r|.
%
Instead, we simply pipe the generation to the inner |HRep|, calling itself on
any recursive generation call that might appear inside:

@@

This way we obtain a concrete recursive generator for each representation |f|
that we define.
%
Then, we can define a random generator for our target data type simply by
generating a random value of our chosen representation, and transforming it back
to our target by the means of the evaluation relation:

@@

So far we have introduced the machinery required to represent the random
generation of a target data type considering only the structure encoded on its
definition.
%
However, this approach can now be easily extended to encode different sources of
structured information.
%
We proceed to introduce two extensions that help to address the problematic
testing scenarios presented in the previous section.

@@

The first pattern of |foo| contains two pattern variables (|x| and |y|) of type
|Exp| that are abstracted in the parametric fields of |Mk_foo_1|.
%
Similarly, the second pattern of |foo| contains the pattern variable |x| of type
|Int| represented by the first field of |Mk_foo_2|, along with the pattern
variable |y| of type |Exp| that also is abstracted in the second field of
|Mk_foo_2|.

To transform these representations back to our target data type, we simply
expand them it into the concrete value represented by the original pattern,
evaluating its fields back to |Exp| as well:
%
\begin{align*}
  &\evalrep{|Mk_foo_1|\ x\ y}{Exp} = \\
  &\ \ |Add (Add eval_x_Exp (Val 50)) (Add (Val 25) eval_y_Exp)|\\
  &\evalrep{|Mk_foo_2|\ x\ y}{Exp} = \\
  &\ \ |Mul (Val 50) (Mul (Val x) eval_y_Exp)|
\end{align*}

And the random generation of pattern |HRep|s is defined in the same way as we
did before for data constructors |HRep|s:

@@

Note that, by doing this, the generation process inherits any patology that the
functions we use to generate values might have.
%
For instance, if the function |square| is non-terminating for some inputs, our
generation process could suffer from this as well.
%
Furthermore, the generation procedure for abstract interface representations
follows the same pattern as before:

@@

Finally, we can put all the derived machinery together into a generation
specification |Exp_S|, assigning (possibly) different generation frequencies to
each individual |HRep| we combine:

\begin{code}
type Exp_S  =       HRep_Exp  otimes 4
            oplus'  HRep_foo  otimes 2
            oplus'  HRep_M
\end{code}

@@

We want to remark that, for space reasons, we were only able to introduce the
representation of a rather simple target data type.
%
In practice, this reasoning can be extended to mutually recursive and parametric
types as well.

Overall, we believe that this approach has some advantages over the usual
type-driven derivation of random generators:
%
\begin{itemize}
\item \textbf{Composability:} we can combine different atomic representations
  using different structure information sources depending on what property or
  sub-system we need to verify using random values.
  %
\item \textbf{Extensibility:} the developer can derive representations for new
  sources of structure information and combine them with the existing ones
  simply by adding them to the exiting representation of the target data type.
  %
\item \textbf{Predictability:} using branching processes theory, it is possible
  to predict the average distribution of generated values in terms of number of
  constructors.
  %
  This prediction is completely modular, and can be obtained for any composite
  representation obtained using the automatically derived |HRep|s and the
  provided combinators.
\end{itemize}

\todo[inline]{Maybe we can show an example random value from the representation
  and its corresponding target value }

@@

% \begin{figure}[b]
%   \centering
%   \input{tikz/val.tex}
%   \caption{Higher level representation of the data type |Exp|, defined using
%     structural information from the function |foo| and the abstract interface of
%     the module |M|.}
%   \label{fig:hrep}
% \end{figure}
